---
title: "dplyrr - Tools for improvement of experience on dplyr with databases"
author: "Koji MAKIYAMA"
output:
  html_document:
    keep_md: yes
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE)
```

## Overview

`dplyr` is the most powerful package for data handling in R, and it has also the ability of working with databases([See Vignette](http://cran.rstudio.com/web/packages/dplyr/vignettes/databases.html)).  
But the functionalities of dealing with databases in `dplyr` is developing yet.

Now, I'm trying to make `dplyr` with databases more comfortable by using some functions.  
For that purpose, I've created `dplyrr` package.

`dplyrr` has below functions:

- `load_tbls()` : Easy to load table objects for all tables in a database.
- `case_cut()` : Easy to create a case statement by using the grammar like the `base::cut()`.
- `filter()` : Improved `filter()` for `tbl_sql` which adds parentheses appropriately.
- `moving_average()` : Compute moving average for PostgreSQL.

`dplyrr` is going to have below functions:

- Compute `first_value` for PostgreSQL.

## How to install

```{r, eval=FALSE}
install.packages("devtools") # if you have not installed "devtools" package
devtools::install_github("hoxo-m/dplyrr")
```

## Common functions for all databases

For illustration, we use a database file: "my_db.sqlite3".  
If you want to trace the codes below, you should first create the databese file.

```{r, eval=FALSE}
library(dplyrr)
library(nycflights13)

db <- src_sqlite("my_db.sqlite3", create = TRUE)
copy_nycflights13(db)
```

### `load_tbls()`

Usually, when we use a database with `dplyr`, we first create database object, and we can see the tables in the databese by `show()`.

```{r}
library(dplyrr)
# Create database object
db <- src_sqlite("my_db.sqlite3")
show(db)
```

Next, we create table objects for pulling data in some tables in the database.

```{r, eval=FALSE}
airlines_tbl <- tbl(db, "airlines")
airports_tbl <- tbl(db, "airports")
flights_tbl <- tbl(db, "flights")
planes_tbl <- tbl(db, "planes")
weather_tbl <- tbl(db, "weather")
```

Typing this code is really a bore!

If you want to create table objects for **all tables in the database**, you can use `load_tbls()`.

```{r message=TRUE}
load_tbls(db)
```

Check the created table objects.

```{r}
ls(pattern = "_tbl$")
glimpse(airlines_tbl)
```

### `case_cut()`

If you want to write case statement with like `cut()`, you can `case_cut()` function.

For example, there is `air_time` column in the database.

```{r}
db <- src_sqlite("my_db.sqlite3")
flights_tbl <- tbl(db, "flights")
q <- flights_tbl %>% select(air_time)
air_time <- q %>% collect
head(air_time, 3)
```

If you want to group the `air_time` by break points `c(0, 80, 120, 190, 900)`, you think you must write next code.

```{r}
q <- flights_tbl %>% 
  select(air_time) %>%
  mutate(air_time_cut=if(air_time > 0 && air_time <= 80) "(0,80]"
         else if(air_time > 80 && air_time <= 120) "(80,120]"
         else if(air_time > 120 && air_time <= 190) "(120,190]"
         else if(air_time > 190 && air_time <= 900) "(190,900]")
air_time_with_cut <- q %>% collect
head(air_time_with_cut, 3)
```

When the break points increase, you are going to be tired to write more lines.

Using `case_cut()` function, it can be easily.

```{r}
q <- flights_tbl %>% 
  select(air_time) %>%
  case_cut(air_time_cut=air_time, breaks=c(0, 80, 120, 190, 900))
air_time_with_cut <- q %>% collect
head(air_time_with_cut, 3)
```

The `case_cut()` has more arguments such as `labels` coming from `base::cut()`.

### Improved `filter()`

```{r echo=FALSE}
detach("package:dplyrr", unload = TRUE)
detach("package:dplyr", unload = TRUE)
```

If you use `dplyr` with databases in pure mind, you can encounter the unintended action like below.

```{r}
library(dplyr)

db <- src_sqlite("my_db.sqlite3")
flights_tbl <- tbl(db, "flights")
q <- flights_tbl %>%
  select(month, air_time) %>%
  filter(month == 1) %>%
  filter(air_time > 200 || air_time < 100)
q$query
```

Did you expect the where clause to be that?

If you use `dplyrr`, it becomes natural by adding parentheses.

```{r}
library(dplyrr)

db <- src_sqlite("my_db.sqlite3")
flights_tbl <- tbl(db, "flights")
q <- flights_tbl %>%
  select(month, air_time) %>%
  filter(month == 1) %>%
  filter(air_time > 200 || air_time < 100)
q$query
```

## Functions for PostgreSQL

### `moving_average()`

```{r}
db <- src_postgres()
account_tbl <- tbl(db, "account")
q <- account_tbl %>% 
  filter(between(register_datetime, "2010-09-01", "2010-10-01")) %>%
  count(date=date_trunc("days", register_datetime)) %>%
  rename(account_num=n) %>%
  arrange(date) %>%
  moving_average(moving_average_3days=account_num(1), moving_average_5days=account_num(2))
data <- q %>% collect
data
```

